Classical Planning with Simulations: \\ Results on the Atari Video Games

### 2015-02-04 ijcai15-paper/paper.tex
First iteration IJCAI-15

 \documentclass[letterpaper]{article}
\usepackage{aaai}
\usepackage{times}
\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\frenchspacing
\setlength{\pdfpagewidth}{8.5in}
\setlength{\pdfpageheight}{11in}

@@

 \begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
\pdfinfo{
/Title (Classical Planning Algorithms on the Atari Video Games)
/Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

\title{Classical Planning Algorithms ¡ on the Atari Video Games}

% \author{\textbf{ID:} 52 \ \ ; \ \ \textbf{Keywords:} classical planning, video games}

 \author{Nir Lipovetzky \\
       The University of Melbourne \\
    Melbourne, Australia\\
%     {\normalsize\url{nir.lipovetzky@unimelb.edu.au}}
%%% footnote doesn't look nice ; this looks better
%% although can't use Nir's unimelb address that is too long :-)
%% 
    {\small\url{nirlipo@gmail.com}}
    \And
    Miquel Ramirez \\
    Australian National University \\
    Canberra, Australia\\
   {\small\url{miquel.ramirez@gmail.com}}
    \And
    Hector Geffner \\
    ICREA \&  Universitat Pompeu Fabra \\
    Barcelona, SPAIN \\
    {\small\url{hector.geffner@upf.edu}}
 %%%%% \thanks{firstname.lastname}
}

%AAAI Press\\
%Association for the Advancement of Artificial Intelligence\\
%2275 East Bayshore Road, Suite 160\\
% Palo Alto, California 94303\\}

@@

%% Pending: randomization, discount factor, iw11 ..
%% iw2 multivalued expands up to 256^2 n^nodes, while iw2 boolean 16^2 n^2 nodes; can iw2 boolean beat iw1 multivalued?
%% Important. optimization, need Dikjstra por postprocessing .. see emails

\begin{abstract}
\begin{quote}
The Atari 2600 games supported in  the Arcade Learning Environment~\cite{bellemare:jair2013}
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  for selecting actions  for  two reasons: first,  no  compact PDDL-model of  the games is given,
and  more importantly,  the action effects and goals are  not known \emph{a priori}. Moreover, in these games there is usually
no set of goals to be achieved but rewards to be collected. These features do not preclude the use of   classical  algorithms like
breadth-first search or Dijkstra's algorithm,  but    these methods are not effective over  large state spaces. We thus turn to a
different class of classical planning algorithms  introduced recently  that perform a \emph{structured exploration} of the state space; namely,
like breadth-first search and Dijkstra's  algorithm they are ``blind'' and hence  do not require prior knowledge of state transitions,
costs (rewards) or goals, and yet, like  heuristic search algorithms, they have been shown to be   effective for solving   problems over huge
state spaces. The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
$\IW(k)$ where  $\IW(i)$ is  a breadth-first search in which a state is  pruned when it is not the
first state in the search  to make true some subset of $i$ atoms. 
The  empirical results over \numgames\ games suggest that the performance 
of \IW\ with the  $k$ parameter fixed to $1$, i.e., $\IW(1)$, is at the level of the state of the art represented by 
\UCT.  A simple best-first variation  of \IW\  that combines exploration and exploitation proves
to be very competitive as well. 
\end{quote}
\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with state-of-the-art classical planners \cite{geffner:book}.
This is because there is  no compact PDDL-like encoding of the domain, and more
importantly,  the goal to be achieved in each game  is not given. Indeed, there are
often no goals  but rewards $r(a,s)$ that  depend on the action $a$ and
the state $s$ in  which  the action is performed, and these rewards are not known.\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}
Thus no model of the  goals  or the  rewards can be used to bias the search.

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line.
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planners: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set)
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is
complete, while searching the state space in a way that makes use of the
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.

While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$,
the procedure $\IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$.
Elsewhere, it has been shown that  \IW\ exhibits   state-of-the-art performance on most existing planning benchmark domains
when goals are restricted to single atoms  where \IW\ provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a blocks world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that $\IW(2)$ will
not only be complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for \emph{any} instance as long as the goal is atomic.
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built,
simple extensions of the basic \IW\ procedure have been developed  such as Serialized \IW, where \IW\ is
used to achieve the goals one at at time following a simple goal ordering  \cite{nir:ecai2012,nir:ecai2014}.
The extension of \IW\ for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.

@@

The paper is organized as follows. We review the iterated width  algorithm and where it comes from,
look at  simple   variations of the algorithm that appear to be convenient  for the Atari
games, present and analyze the  experimental results, and discuss related and future work.

@@

For classical planning, the termination condition is the achievement of the goal. In the on-line setting,
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW \ is the path  that has a  maximum accumulated reward.
The accumulated reward $R(s)$  of a state $s$ reached in an iteration of \IW \ is determined by the  unique parent state $s'$
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected to be done next is  the first action along  such a path.

@@

\begin{table}[t!]
\begin{center}
%\resizebox{3in}{!}{
{\scriptsize
% \begin{tabular}{r@{}lrrrr@{}}
\begin{tabular}{llcccc}
%% table header
\hline\\[-1.3ex]
%\toprule
\# & \multicolumn{1}{c}{Domain} & \multicolumn{1}{c}{I} & \multicolumn{1}{c}{$\IW(1)$} & \multicolumn{1}{c}{$\IW(2)$} & \multicolumn{1}{c}{Neither} \\[.2ex]

\hline\\[-1.3ex]
%\hline
%% table body
1. & 8puzzle	&400	&55\%	&45\%	&0\%\\
2. & Barman	&232	&9\%	&0\%	&91\%\\
3. & Blocks World	&598	&26\%	&74\%	&0\%\\
4. &Cybersecure	&86	&65\%	&0\%	&35\%\\
% 5. &Depots	&189	&11\%	&66\%	&23\%\\
% 6. & Driver	&259	&45\%	&55\%	&0\%\\
% 7. & Elevators	&510	&0\%	&100\%	&0\%\\
% 8. & Ferry	&650	&36\%	&64\%	&0\%\\
% 9. & Floortile	&538	&96\%	&4\%	&0\%\\
% 10. & Freecell	&76	&8\%	&92\%	&0\%\\
% 11. & Grid	&19	&5\%	&84\%	&11\%\\
% 12. & Gripper	&1275	&0\%	&100\%	&0\%\\
% 13. & Logistics	&249	&18\%	&82\%	&0\%\\
% 14. & Miconic	&650	&0\%	&100\%	&0\%\\
% 15. & Mprime	&43	&5\%	&95\%	&0\%\\
% 16. & Mystery	&30	&7\%	&93\%	&0\%\\
% 17. & NoMystery	&210	&0\%	&100\%	&0\%\\
% 18. & OpenStacks	&630	&0\%	&0\%	&100\%\\
% 19. & OpenStacksIPC6	&1230	&5\%	&16\%	&79\%\\
\ldots & \ldots & \ldots & \ldots & \ldots & \\
20. & ParcPrinter	&975	&85\%	&15\%	&0\%\\
21. & Parking	&540	&77\%	&23\%	&0\%\\
22. & Pegsol	&964	&92\%	&8\%	&0\%\\
23. & Pipes-NonTan	&259	&44\%	&56\%	&0\%\\
24. & Pipes-Tan	&369	&59\%	&37\%	&3\%\\
25. & PSRsmall	&316	&92\%	&0\%	&8\%\\
26. & Rovers	&488	&47\%	&53\%	&0\%\\
27. & Satellite	&308	&11\%	&89\%	&0\%\\
28. & Scanalyzer	&624	&100\%	&0\%	&0\%\\
29. & Sokoban	&153	&37\%	&36\%	&27\%\\
30. & Storage	&240	&100\%	&0\%	&0\%\\
31. & Tidybot	&84	&12\%	&39\%	&49\%\\
32. & Tpp	&315	&0\%	&92\%	&8\%\\
33. & Transport	&330	&0\%	&100\%	&0\%\\
34. & Trucks	&345	&0\%	&100\%	&0\%\\
35. & Visitall	&21859	&100\%	&0\%	&0\%\\
36. & Woodworking	&1659	&100\%	&0\%	&0\%\\
37. & Zeno	&219	&21\%	&79\%	&0\%\\[.2ex]
%\hline
\hline\\[-1.3ex]
& Total/Avgs&37921	&37.0\%	&51.3\%	&11.7\%\\
\hline
\end{tabular}
}

\vskip .1cm

{\scriptsize
\begin{tabular}{@{}ccccc@{}}
\hline\\[-1.3ex]
\multicolumn{1}{c}{\# Instances}  &\multicolumn{1}{c}{\IW} & \multicolumn{1}{c}{\ID} & \multicolumn{1}{c}{\BRFS} &\multicolumn{1}{c}{\textit{GBFS +} $h_{add}$} \\[.2ex]
\hline\\[-1.3ex]
37921  &34627 &9010	&8762 &34849 \\
\hline
%\hline
\end{tabular}
}

\end{center}

\caption{\small
\emph{Top:} Number of classical planning instances per domain  and percentages solved by $\IW(1)$, $\IW(2)$, or neither.
Problems  obtained by splitting benchmarks  with $N$ atomic goals into $N$ problems with atomic goals.
\emph{Bottom:} Number of instances solved by \IW\ in comparison with  Iterative Deepening (\ID),
Breadth-First Search (\BRFS),   and Greedy Best First Search (\textit{GBFS}) guided by  additive heuristic.
Time and  memory outs  after 30 minutes or 2 GB. Table from \cite{nir:ecai2012}.
}
\label{table1}
\end{table}
%%%%
%%%

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (\ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search in a different way. This structured exploration  has proved to be very
effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.}
Table~\ref{table1} from \cite{nir:ecai2012} shows
the percentage of instances that are  solved by  the first iteration of \IW, i.e.  $\IW(1)$, by the second iteration
$\IW(2)$, and by neither one.  These are instances that  have been obtained from the existing benchmarks
by splitting problems with $N$ atomic goals, into $N$ problems with one atomic goal each. As the table shows,
$37\%$ of the 37921 instances are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$. Since $\IW(k)$
runs in time that is exponential in $k$, this mean that almost $90\%$ of the 37,921 instances are  solved in time that
is either linear or quadratic in the number of problem variables, which in these encoding are all \emph{boolean}.
Furthermore, when the performance of \IW\   is compared with  breadth-first search and iterative deepening,
on the one hand,  and  with a  Greedy Best First Search guided by the additive heuristic $h_{add}$
\cite{bonet:aij-hsp} on the other (this algorithm is similar
to the one used by  the FF planner when hill climbing search fails \cite{hoffmann:ff}),
it turns out that ``blind'' \IW\ solves as many problems as the informed search, 34,627 vs. 34,849, far ahead
of the other blind algorithms \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. This is shown in the bottom
part of  Table~\ref{table1}. Moreover, \IW\ is  faster and results in shorter plans than a heuristic search algorithm
\cite{nir:ecai2012}.

The min $k$ value for which $\IW(k)$ solves a problem  is bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{k+1}$ is
at distance 1 from the previous one $t_k$ when $t_k$ has been reached following the ``trail'',
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_k$ can be reached in less than $k$ steps.

@@

@@

 \caption{\small Performance that results form various  lookahead algorithms in \numgames\  Atari games. The algorithms, \BRFS, $\IW(1)$, \BFS, and \UCT, are evaluated
 over $10$ runs (episodes) for each game. 
%%% Averages scores  over 5 runs shown with ``random'' seeds $0$ to $4$,  
%%% this does not apply to UCT and BRFS if taken from JAIR ..
%% 
The maximum episode duration is  $18,000$  frames and every algorithm is limited to a lookahead budget of 150,000 simulated frames.
% , except for UCT limited to 500 rollouts, 300 frames deep instead. 
Figures for \BRFS\ and \UCT\ taken from \cite{bellemare:jair2013}.
Average  CPU times per action in seconds, rounded to nearest integer, shown for  $\IW(1)$ and \BFS. 
Numbers  in bold show best performer in terms of average score, while numbers shaded in light grey show scores  that are better than \UCT's. 
Bottom part of the table shows pairwise comparisons among the algorithms. 
% Entries shown as ``--'' are still running at the time of submission. Each run can take up to 24 hours of CPU time. Eventually
% all 55 games from  \cite{bellemare:jair2013}  to be included.
}
\label{performance}

\end{table*}

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to $\IW(1)$, \BFS, and \UCT. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its pruning does not always use
the full budget and takes less time per action on average.

@@

\section{Summary and Future Work}

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. This level of play 
is the result of a structured exploration of the state space that manages to combine 
the scope of blind search algorithms with the efficiency of heuristic search methods. 
There are several  issues worth studying, we  mention two of them.
Regarding planning, the results show the potential of width-based algorithms
for a broader class of planning problems. As shown, indeed, these algorithms
do not require PDDL encodings and can deal with state successor functions
encoded \emph{procedurally} (as in simulators) as long as the structure of the states 
is known. 
%% This suggests to us that the algorithms discussed in the 
%% present paper could be applied in a broad variety of settings where only on-line 
%% action selection is feasible or suitable.

Concerning the Atari games, there are two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not available to the human player. And second, they do not get 
better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
state of the screen pixels as opposed to the RAM state, and the potential uses of the new ideas for learning,
such the use of the  novelty measures for shaping rewards.

@@

\noindent \textbf{Acknowledgments.} We thank the people involved in
the creation and maintenance of the Arcade Learning Environment that
have made this work possible. The research by Nir Lipovetzky is partially
funded by the Australian Research Council Linkage grant LP11010015.

@@

\bibliographystyle{aaai}
\bibliography{control}

>>>>

% \documentclass[letterpaper]{article}

\documentclass{article}
% The file ijcai15.sty is the style file for IJCAI-15 (same as ijcai07.sty).
\usepackage{ijcai15}

% Use the postscript times font!
\usepackage{times}

% \usepackage{aaai}
% \usepackage{times}

\usepackage{helvet}
\usepackage{courier}
\usepackage{url}
\frenchspacing
% \setlength{\pdfpagewidth}{8.5in}
% \setlength{\pdfpageheight}{11in}

@@

\begin{document}
% The file aaai.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%
% \pdfinfo{
% /Title (Classical Planning Algorithms on the Atari Video Games)
% /Author (Nir Lipovetzky, Miquel Ramirez, Hector Geffner)}

@@

\title{Classical Planning with Simulations: \\ Results on the Atari Video Games}

\author{\textbf{ID:} \ldots  \ \ ; \ \ \textbf{Keywords:} classical planning, search, video games}

% \author{Nir Lipovetzky \\
%        The University of Melbourne \\
%     Melbourne, Australia\\
%     {\normalsize\url{nir.lipovetzky@unimelb.edu.au}}
%%% footnote doesn't look nice ; this looks better
%% although can't use Nir's unimelb address that is too long :-)
%% 
%    {\small\url{nirlipo@gmail.com}}
%    \And
%    Miquel Ramirez \\
%    Australian National University \\
%    Canberra, Australia\\
%   {\small\url{miquel.ramirez@gmail.com}}
%    \And
%    Hector Geffner \\
%    ICREA \&  Universitat Pompeu Fabra \\
%    Barcelona, SPAIN \\
%    {\small\url{hector.geffner@upf.edu}}
 %%%%% \thanks{firstname.lastname}
%}

@@

\begin{abstract}
The Atari 2600 games supported in  the Arcade Learning Environment~\cite{bellemare:jair2013}
all feature a known initial (RAM) state and actions that have deterministic effects. Classical planners, however,
cannot be used  off-the-shelf as there is no  compact PDDL-model of  the games, and action effects and goals are  not
known \emph{a priori}. Indeed, there are no explicit goals, and the planner must select actions on-line while interacting 
with a simulator that returns successor states and rewards. None of this precludes  the use of  blind lookahead algorithms 
for action selection like breadth-first search or  Dijkstra's yet such methods are not effective over large  state spaces. 
We thus turn to a different class of  planning methods  introduced recently  that  have been shown to be  effective for solving
large planning problems but which do not require  prior knowledge of state transitions and costs (rewards) or goals.
The simplest such algorithm, called Iterated Width or \IW, consists of  a sequence of calls  $\IW(1)$, $\IW(2)$, \ldots,
where  $\IW(k)$ is  a breadth-first search in which a state is  pruned when it is not the first state in the  search  to make true 
a subset of $k$ atoms. The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.
\end{abstract}

@@

While all  these games feature a known initial (RAM) state and actions that
have deterministic effects, the problem of selecting the next action
to be done cannot be addressed with off-the-shelf classical planners 
\cite{planning:book,geffner:book}. This is because there is  no compact PDDL-like encoding of the domain and 
the goal to be achieved in each game  is not given. Indeed, there are no goals but rewards, and 
the planner must select actions on-line while interacting with a simulator that just
returns successor states and rewards.
%\footnote{Actually, in the Atari games, the rewards $r(a,s)$ depend only
% on the state $f(a,s)$ that results from doing action $a$ in the state $s$.}

@@

The limitation of the   basic blind search  methods is that they are not effective  over large  spaces, neither
for solving problems off-line, nor for guiding  a lookahead search for solving problems on-line.
In this work, we thus turn to a recent class of  planning algorithms that combine the scope of blind search
methods with the performance of state-of-the-art classical planners: namely, like ``blind'' search algorithms
they do not require prior knowledge of state transitions, costs, or goals, and yet like heuristic algorithms
they manage to search large  state spaces effectively. The basic algorithm in this class is
called \IW\ for Iterated Width search \cite{nir:ecai2012}. \IW\ consists of a sequence of calls  $\IW(1)$, $\IW(2)$, .., $\IW(k)$, where
$\IW(i)$ is  a  standard breadth-first search where states are pruned right away when they fail to make true some new tuple (set)
of at most $i$ atoms. Namely,   $\IW(1)$ is a breadth-first search that keeps  a state only if  the state is   the first  one
in the search to make some atom  true; $\IW(2)$ keeps a state  only if the state is  the first one to  make  a pair of atoms true, and
so on.  Like plain breadth-first  and iterative deepening searches,  \IW\  is
complete, while searching the state space in a way that makes use of the
\emph{structure of states} given by the values of a finite set of state variables.
In the Atari games, the (RAM) state is given by a vector of 128 bytes, which we associate with 128 variables
$X_i$, $i=1, \ldots, 128$, each of which may take up to $256$ values $x_j$. A state $s$ makes an atom $X_i=x_{j}$
true when the value of the $i$-th byte in the state vector $s$ is $x_j$.
The  empirical results over \numgames\ Atari games show that  $\IW(1)$ performs at the level of UCT, 
the state-of-the-art planning method in this domain, and suggest the potential of width-based methods for planning with
simulators when factored, compact action models are not available.

\Omit{
While a normal, complete breadth-first search runs in time that is exponential in the total number of variables $X_i$,
the procedure $\IW(k)$ runs in time that is exponential in the $k$ parameter, $1 \leq k \leq n$.
Elsewhere, it has been shown that  \IW\ exhibits   state-of-the-art performance on most existing planning benchmark domains
when goals are restricted to single atoms  where \IW\ provably runs in low polynomial time in the number of
variables \cite{nir:ecai2012}. For example, in a blocks world instance with any number of blocks and any initial
configuration,  the goal of  placing a given  block on top of another will be achievable in  quadratic time
in the number of blocks as any such instance can be shown to  have \emph{width} 2. This means  that $\IW(2)$ will
not only be complete then, but also optimal. This is also true for most other benchmark domains in classical planning
that turn out to  have small, bounded widths for \emph{any} instance as long as the goal is atomic.
For dealing with the benchmarks where goals are  not single atoms, e.g., where a given \emph{tower} of blocks needs to be built,
simple extensions of the basic \IW\ procedure have been developed  such as Serialized \IW, where \IW\ is
used to achieve the goals one at at time following a simple goal ordering  \cite{nir:ecai2012,nir:ecai2014}.
The extension of \IW\ for problems where there is no goal but an
additive reward measure  to be optimized, like in the Atari games, is direct.
}

@@

The paper is organized as follows. We review the iterated width  algorithm and its properties,
look at the    variations of the algorithm that we used in the Atari games, present and analyze the  
experimental results, and discuss related and future work.

@@

For classical planning, the termination condition is the achievement of the goal. In the \emph{on-line setting,}
as in the Atari games, the termination condition is given by a time window or a maximum number of generated nodes.
The  best path found by \IW \ is then the  path  that has a  maximum accumulated reward. The accumulated reward $R(s)$  
of a state $s$ reached in an iteration of \IW \ is determined by the  unique parent state $s'$
and action $a$ leading to $s$ from $s'$ as $R(s) = R(s') + r(a,s')$. The best state is the state $s$ with maximum reward
$R(s)$ generated but not pruned by \IW, and the best path is the one that leads to the state $s$ from the current state.
The action selected  in the on-line setting  is  the first action along  such a path. This action is then executed
and the process repeats from the resulting state.

@@

@@

\IW\ is a systematic and complete blind-search algorithm like breadth-first search (\BRFS) and iterative deepening (\ID),
but   unlike these algorithms, it uses the factored representation of the states in terms of variables to structure the
search. This structured exploration  has proved to be very effective over classical planning benchmark domains  when goals are  single atoms.\footnote{
Any conjunctive goal can be mapped into a single dummy atomic goal by adding an action that
achieves the dummy goal and that has the original conjunctive goal as a precondition.
Yet, this changes the definition of the domain.} For example, $37\%$ of the 37,921 problems
considered  in \cite{nir:ecai2012} are solved by $\IW(1)$ while $51.3\%$ are  solved by $\IW(2)$.
These are instances obtained from 37 benchmark domains by splitting problems with $N$ atomic
goals into $N$ problems with one atomic goal each.  Since $\IW(k)$ runs in time that is exponential in $k$, 
this mean that almost $90\%$ of the 37,921 instances are  solved in time that is either linear or quadratic in the number of problem variables, 
which in such encodings are all \emph{boolean}. Furthermore, when the performance of \IW\   is compared with  
a  Greedy Best First Search guided by the additive heuristic $h_{add}$, it turns out that ``blind'' \IW\ solves as many problems as the informed 
search, 34,627 vs. 34,849, far ahead of  other blind search  algorithms like \BRFS\ and \ID\  that solve 9,010 and 8,762 problems each. 
Moreover, \IW\ is  faster and results in shorter plans than in the heuristic search.

The min $k$ value for which $\IW(k)$ solves a problem  is indeed bounded and small in most of
these instances. This is  actually no  accident and  has a \emph{theoretical explanation.} Lipovetzky and Geffner define
a structural parameter called the problem \emph{width} and show that for many of these domains,
\emph{any} solvable instance with atomic goals will  have a bounded and small width that is independent of the number
of variables and states in the problem. The min value $k$ for which the iteration  $\IW(k)$ solves
the problem cannot exceed the problem width, so  the algorithm \emph{\IW\ runs in time and space
that are exponential in the problem width.}

@@

\Omit{
One way to understand this definition is that the problem  width $w(P)$ is at most  $i$ if  there is a
``trail of  stepping stones''  $t_0, t_1, \ldots, t_n$ to reach the problem goal such that
A)~these ``stepping stones''  contain at most $i$ atoms each, B)~each  stepping stone $t_{k+1}$ is
at distance 1 from the previous one $t_k$ when $t_k$ has been reached following the ``trail'',
C)~the ``trail''  preserves ``optimality''; i.e., no  tuple $t_k$ can be reached in less than $k$ steps.
}

@@

We tested \IW(1) and \BFS\ over \numgames\  of the $55$ different games 
considered in \cite{bellemare:jair2013}, from now on  abbreviated as BNVB.\footnote{
We left out  \textsc{Skiing} as the reported figures apparently use a  different reward structure.}
%% check
% \footnote{
% Entries marked as  ``--'' in the table are   still running at the time of submission. 
% Each run can take up to $24$ hours of CPU time and there are $10$ runs per game and algorithm. 
% Eventually all of  the  $55$ games will  be included in the evaluation.}
The two  algorithms were used to play the  games in the \emph{on-line planning
setting} supported by ALE where we will compare them with the planning algorithms 
considered by BNVB; namely, breadth-first search and \UCT. ALE supports also a \emph{learning setting} 
where the goal is to learn controllers that map  states into actions without
doing any lookahead. Algorithms across  the two settings are thus not directly comparable
as they compute different things. Learning controllers appears as a more challenging problem
and it is thus not surprising that planning algorithms like UCT tend to achieve a higher score than 
learning algorithms. In addition, the learning algorithms reported by BNVB tend to use 
the state of the screen pixels, while the planning algorithms,  use the state
of the RAM memory. It is not clear however whether one input representation is
more challenging than the other; for the learning algorithms, BNVB mention
that the results tend to be better for the screen inputs.

Table~\ref{performance} shows the performance of $\IW(1)$ and \BFS\
in comparison  with breadth-first search (\BRFS) and \UCT.
Videos of selected games played by $\IW(1)$, \BFS, and \UCT\ can be 
seen in Youtube.\footnote{\url{http://www.youtube.com/playlist?list=PLXpQcXUQ_CwenUazUivhXyYvjuS6KQOI0}.}
The discount factor used by  all the algorithms is $\gamma=0.995$.
The scores reported for \BRFS\ and \UCT\ are taken from BNVB.  
Our experimental setup  follows theirs except that a maximum budget of  $150,000$ simulated frames 
is applied to $\IW(1)$, \BFS, and \UCT. \UCT\ uses this budget by running
 $500$   rollouts of depth $300$. % \footnote{In \UCT,  $500$ base policy rollouts
% of depth $300$ is equivalent to a budget of $150,000$ simulated frames.}.
The bound on the number of simulated frames is like a bound on
lookahead time, as most of the time in the lookahead  is spent
in calls to the emulator for computing the next RAM state. 
This is why the average time per action is similar to all the algorithms
except $\IW(1)$, that due to its pruning does not always use
the full budget and takes less time per action on average.

@@

 \caption{\small Performance that results form various  lookahead algorithms in \numgames\  Atari games. The algorithms, \BRFS, $\IW(1)$, \BFS, and \UCT, are evaluated
 over $10$ runs (episodes) for each game. 
The maximum episode duration is  $18,000$  frames and every algorithm is limited to a lookahead budget of 150,000 simulated frames.
Figures for \BRFS\ and \UCT\ taken from Bellemare et al. Average  CPU times per action in seconds, rounded to nearest integer, shown for  $\IW(1)$ and \BFS. 
Numbers  in bold show best performer in terms of average score, while numbers shaded in light grey show scores  that are better than \UCT's. 
Bottom part of the table shows pairwise comparisons among the algorithms.}
\label{performance}
\end{table*}

@@

\section{Summary}

We have shown experimentally that width-based algorithms like $\IW(1)$ and \BFS\
that originate in work in classical planning,  can be used to play  the Atari video
games where they  achieve  state-of-the-art performance. The  results also  suggest 
more generally the potential of width-based methods for planning with simulators when factored, 
compact action models are not available. In this sense, the scope of these planning methods 
is broader than those of heuristic-search planning methods that require propositional
encodings of actions and goals, and with suitable extensions, 
may potentially  approach the scope of MCTS-methods like UCT that work on simulators as well.

\newpage

%% This suggests to us that the algorithms discussed in the 
%% present paper could be applied in a broad variety of settings where only on-line 
%% action selection is feasible or suitable.

\Omit{
Concerning the Atari games, there are two problems in all the planning methods
discussed. First, by using the RAM state of the Atari console one can consider
that they are ``cheating'', as they use information that is not available to the human player. And second, they do not get 
better with experience. In the future, we would like to explore 
whether variations of these algorithms  can be used to play from the information conveyed by the  
state of the screen pixels as opposed to the RAM state, and the potential uses of the new ideas for learning,
such the use of the  novelty measures for shaping rewards.

@@

\noindent \textbf{Acknowledgments.} We thank the people involved in
the creation and maintenance of the Arcade Learning Environment that
have made this work possible. The research by Nir Lipovetzky is partially
funded by the Australian Research Council Linkage grant LP11010015.
}

@@

\bibliographystyle{named}
\bibliography{control}
